cardinalR: Generating interesting high-dimensional data structures

### 2025-09-08 paper-cardinalR.tex
Rendered new one

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The \texttt{geozoo} package (\citet{barret2016}) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is \texttt{snedata} (\citet{james2025}), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, \texttt{mlbench} (\citet{friedrich2024}) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, \texttt{splatter} (\citet{luke2017}) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression.

There is an opportunity to improve the generation of high-dimensional data structures by integrating geometric principles with advanced noise control and customizable clustering. The \texttt{geozoo} package provides a strong foundation but could be enhanced to support high-dimensional extensions with controlled noise and user-defined parameters for clustering. Similarly, while \texttt{snedata} focuses on abstract datasets for dimensionality reduction, adding features for generating high-dimensional data from geometric layouts would enhance its usability. The \texttt{mlbench} package could also benefit from allowing users to create datasets with specific geometric structures and noise profiles. Additionally, although \texttt{splatter} specializes in biological data simulation, it could be expanded to offer a broader framework for generating diverse geometric structures across dimensions, enabling detailed control over noise and clustering. Addressing these areas could lead to more robust high-dimensional data generation tools.

To address this gap, this paper introduces the \texttt{cardinalR} R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. \texttt{cardinalR} offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By providing these features, \texttt{cardinalR} provides researchers with a method to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.

The paper is organized as follows. In next section, introduces the implementation of \texttt{cardinalR} package on CRAN and GitHub, including demonstration of the package's key functions. We illustrate how a clustering data structure affect the dimension reductions in \textbf{Application} section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for use of our data collection.

@@

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} c). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\) dimensions contain Gaussian noise.

Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1X_1 + X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are added iteratively. Each new branch \(i\) begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X_{1}\):\(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure: \(X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}\). Here, \(s_i\) is a sampled scale factor (slope-like term) controlling curvature, and \((x_{\text{start}}, y_{\text{start}})\) is the starting point from the existing structure. This construction yields \(k\) spatially connected, nonlinear branches with varying curvature. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

The \texttt{gen\_orglinearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} d). Each branch lies primarily within a distinct \(2\text{-}D\) subspace, while the remaining \(p - 2\) dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the \(\binom{p}{2}\) possible \(2\text{-}D\) combinations. If \(k \leq \binom{p}{2}\), combinations are sampled without replacement. If \(k > \binom{p}{2}\), additional pairs are sampled with replacement to reach \(k\) total branches. Each selected pair \((X_{i1}, X_{i2})\) defines the \(2\text{-}D\) plane for branch \(i\).

For branch \(i\), \(n_i\) points are generated, where \(\sum_{i=1}^k n_i = n\). The structure follows: \(X_{i1} \sim U(0, 2)\), and \(X_{i2} = -s_iX_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5)\), where \(s_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s_i = 1\). When sampling with replacement, \(s_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\). For \(p > 2\), the remaining dimensions contain independent Gaussian noise: \(X_j \sim N(0, 0.1^2)\), for \(j \notin \{i1, i2\}\).

@@

Let \(\binom{p}{2}\) denote the number of unique \(2\text{-}D\) subspace combinations. When \(k \leq \binom{p}{2}\), \(k\) distinct subspace pairs \((X_{i1}, X_{i2})\) are sampled without replacement. Otherwise, combinations are selected with replacement to reach \(k\) total branches. For each branch \(i = 1, \dots, k\), a random scale factor \(s_i\) is used to vary the curvature; If \(k \leq \binom{p}{2}\): \(s_i = 1\), or If \(k > \binom{p}{2}\): \(s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}\).

Each branch contains \(n_i\) points such that \(\sum_{i=1}^k n_i = n\), where the vector \((n_1, \dots, n_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\). Within its assigned subspace \((X_{i1}, X_{i2})\), branch \(i\) is defined by: \(X_{i1} \sim U(0, 2)\), and \(X_{i2} = -s_iX_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)\). This forms a smooth downward-opening parabola in the plane defined by \((X_{i1}, X_{i2})\), with the degree of curvature controlled by \(s_i\). For \(p > 2\), all remaining dimensions are independent Gaussian noise: \(X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}\).

@@

To simulate a cone-shaped structure (Figure \ref{fig:cone-proj}) in arbitrary dimensions, we define a function \texttt{gen\_cone(n,\ p,\ h,\ ratio)}, which creates a high-dimensional cone with options for a sharp or blunted apex, allowing for a dense concentration of points near the tip.

This function generates \(n\) points in a \(p\text{-}D\), where the last dimension (\(X_p\)) represents height along the cone's axis, and the remaining dimensions define a shrinking hyperspherical cross-section as one moves toward the tip. Points along the height axis are drawn from an exponential distribution to increase density near the tip. The \(X_p = h_i \sim \text{Exp}(\lambda = 2/h)\) truncated at \(h\). The effective radius at height \(h_i\) decreases linearly from base to tip, controlled by a shape ratio parameter \(r_i = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}})h_i/h, \quad \text{where } r_{\text{min}} = \text{ratio},\ r_{\text{max}} = 1\).

>>>>

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The \texttt{geozoo} package (\citet{barret2016}) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is \texttt{snedata} (\citet{james2025}), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, \texttt{mlbench} (\citet{friedrich2024}) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, \texttt{splatter} (\citet{luke2017}) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression. While these packages are valuable, their scope is often limited to specific applications or low-dimensional structures.

To address this gap, this paper introduces the \texttt{cardinalR} R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. \texttt{cardinalR} offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By using clearly defined geometric shapes and controllable features such as number of dimensions, sample size; \texttt{cardinalR} allows researchers to generate transparent and interpretable synthetic datasets for evaluating algorithm performance in high-dimensional settings.

By providing these features, \texttt{cardinalR} provides researchers with a method to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.

The paper is organized as follows. In the next section, we introduce the implementation of the \texttt{cardinalR} package on GitHub, including a demonstration of the package's key functions. We illustrate how a clustering data structure affects the dimension reductions in the Application section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for the use of our data collection.

@@

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} c). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\) dimensions contain Gaussian noise.
Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1X_1 + X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are added iteratively. Each new branch \(i\) begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) to ensure connectivity with earlier branches, spans a unit-length interval on \(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure: \(X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, \; s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}\). Here, \(s_i\) is a sampled scale factor (slope-like term) controlling curvature, and \((x_{\text{start}}, y_{\text{start}})\) is the starting point from the existing structure. This construction yields \(k\) spatially connected, nonlinear branches with varying curvature. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

The \texttt{gen\_orglinearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} d). Each branch lies within a distinct \(2\text{-}D\) subspace, while the remaining \(p - 2\) dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the \(\binom{p}{2}\) possible combinations of dimensions. If \(k \leq \binom{p}{2}\), combinations are sampled without replacement. If \(k > \binom{p}{2}\), additional pairs are sampled with replacement to reach \(k\) total branches. Each selected pair \((i_1, i_2)\) defines the \(2\text{-}D\) plane for branch \(i\).

For branch \(i\), \(n_i\) points are generated, where \(\sum_{i=1}^k n_i = n\). The structure follows: \(X_{i_1} \sim U(0, 2)\), and \(X_{i_2} = -s_iX_{i_1} + \epsilon, \quad \epsilon \sim U(0, 0.5)\), where \(s_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s_i = 1\). When sampling with replacement, \(s_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\). For \(p > 2\), the remaining dimensions contain independent Gaussian noise: \(X_j \sim N(0, 0.1^2)\), for \(j \notin \{i_1, i_2\}\).

@@

Let \(\binom{p}{2}\) denote the number of unique \(2\text{-}D\) subspace combinations. When \(k \leq \binom{p}{2}\), \(k\) distinct subspace pairs \((i_1, i_2)\) are sampled without replacement. Otherwise, combinations are selected with replacement to reach \(k\) total branches. For each branch \(i = 1, \dots, k\), a random scale factor \(s_i\) is used to vary the curvature; If \(k \leq \binom{p}{2}\): \(s_i = 1\), or If \(k > \binom{p}{2}\): \(s_i \in \{1, 1.5, \dots, 8\}\).

Each branch contains \(n_i\) points such that \(\sum_{i=1}^k n_i = n\), where the vector \((n_1, \dots, n_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\). Within its assigned subspace, branch \(i\) is defined by: \(X_{i_1} \sim U(0, 2)\), and \(X_{i_2} = -s_iX_{i_1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)\). This forms a smooth downward-opening parabola in the plane defined by \((X_{i_1}, X_{i_2})\), with the degree of curvature controlled by \(s_i\). For \(p > 2\), all remaining dimensions are independent Gaussian noise: \(X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i_1, i_2\}\).

@@

To simulate a cone-shaped structure in arbitrary dimensions (Figure \ref{fig:cone-proj}), we define a function \texttt{gen\_cone(n,\ p,\ h,\ ratio)}, which creates a high-dimensional cone with options for a sharp or blunted apex, allowing for a dense concentration of points near the tip.

This function generates \(n\) points in \(p\text{-}D\), where the last dimension (\(p\)) represents height along the cone's axis, and the remaining dimensions define a shrinking hyperspherical cross-section as one moves toward the tip. Points along the height axis, \(p\), are drawn from an exponential distribution to increase density near the tip. That is, \(X_p = h_i \sim \text{Exp}(\lambda = 2/h)\) truncated at \(h\). The effective radius at height \(h_i\) decreases linearly from base to tip, controlled by a shape ratio parameter \(r_i = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}})h_i/h, \quad \text{where } r_{\text{min}} = \text{ratio},\ r_{\text{max}} = 1\).
